Quantity discrimination, decision-making, and the role of early-life conditions in a lizard

### 2025-03-23 ms.qmd
updates

---
title: "Quantity discrimination, decision-making, and the role of early-life conditions in a lizard "
docx: 
  reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology-and-sociobiology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Abstract
Quantity discrimination affects a range of behaviours that are essential for fitness including social interactions, navigation, and foraging. Animals use different cognitive systems to distinguish between quantities. The Approximate Number System (ANS) estimates, while the Object File System (OFS) tracks discrete items. In reptiles, quantity discrimination remained largely unexplored until recently, and the use of one system over the other can depend on the species or the context in which they are tested. Additionally, early life conditions can influence brain development and cognition, with potential effects on quantity discrimination and decision-making. Here, we tested the ability of the common garden skink (_Lampropholis guichenoti_) to discriminate between quantities and the effects of elevated prenatal GCs and incubation temperature on quantity discrimination and decision-making. We assessed quantity discrimination using a spontaneous choice test with food as a stimulus. We subjected lizards to five two-choice tests each differing in the number of items (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4). Contrary to our predictions, we found no evidence of quantity discrimination in _L. guichenoti_. Furthermore, prenatal conditions did not influence their performance or decision-making. Despite these results, further investigation of quantity discrimination in live-prey-feeding reptiles and the impact of early conditions on other taxa remains an important area of research.

## Significance statement
Discriminating between quantities can be crucial for individual fitness, yet it remains poorly understood in reptiles. In this study, we investigated quantity discrimination in _Lampropholis guichenoti_ using a spontaneous choice test with food as the stimulus. Contrary to our predictions, we found no evidence of quantity discrimination in this species. In addition, we explored the impact of early environmental factors, specifically prenatal corticosterone exposure and incubation temperature, on quantity discrimination and decision-making. However, we did not observe any effect of these early-life conditions on lizards’ behaviour. These results open new avenues for exploring the evolution of quantity discrimination in reptiles and identifying factors influencing cognitive development in this group. Further research should explore alternative methods, such as trained tasks, to better understand the mechanisms underlying quantity discrimination and decision-making in reptiles.

@@

The main two experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. Spontaneous choice tests ascertain if animals can distinguish between two groups of biologically relevant stimuli that differ quantitatively (number, size, volume, etc) [@agrillo2014spontaneous; @cox2016quantity]. In comparison, experiments that employ training procedures evaluate an animal’s ability to learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other, and both can provide different results even in the same species [see @miletto2017quantitative; @miletto2018quantity].

Numerous reasons exist for why methods can lead to different results. First, spontaneous choice tests assume that one of the options presented is more adequate than the other (e.g. the group with more conspecifics or the larger amount of food) when this may not always be the case. For example, choosing groups of conspecifics with less individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [@yee2013costs]. As such, context and motivation can be important drivers of decision-making that can change the way animals discriminate between quantities [@agrillo2014spontaneous]. Second, each method may engage distinct cognitive mechanisms when processing numerical information [@agrillo2014spontaneous]. Different procedures may shape how animals discriminate between quantities, making them more responsive towards certain quantity traits over others. Some tasks may encourage a focus on approximate magnitudes, while others may promote attention to individual items [@agrillo2014spontaneous].

In this regard, researchers propose that two core systems underlie numerical cognition: the approximate number system (ANS) and the object file system (OFS) [@hyde2011two]. The ANS estimates the approximate magnitude of items within a set, with accuracy decreasing as the numerical difference between sets decreases — a principle known as Weber’s law [@hyde2011two]. In contrast, the OFS allows for the precise tracking of discrete items but is thought to be limited to small quantities, typically up to three or four [@hyde2011two]. The engagement of these systems depends not only on the numerical magnitude but also on contextual factors such as the availability of continuous variables (e.g., area, density, movement) and the nature of the task [@stancher2015numerical]. The presence of such continuous cues typically favors ANS-based processing, while in their absence, the OFS is more likely to be employed [@agrillo2014spontaneous]. This often complicates the study of numerosity, as continuous traits must be carefully controlled to ensure the use of the OFS [@agrillo2014spontaneous]. In spontaneous choice tests, for instance, animals may rely on cumulative surface area rather than numerical differences when selecting between two sets of food items [@agrillo2014spontaneous]. To disentangle these cues, researchers often equate the total area of the sets while changing the number of items, forcing individuals to rely on numerical information alone [@stancher2015numerical].

Quantity discrimination studies have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@mehlis2015quantification; @nieder2018evolution]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) and red-footed turtles (_Chelonoidis carbonaria_) can rapidly learn a quantity discrimination task [@soldati2017long; @lin2021superior; @lin2024trained], and Hermann’s tortoises (_Testudo hermanni_) can discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous]. In lizards, the Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. Shinglebacks (_Tiliqua rugosa_) exhibited spontaneous quantity discrimination using both discrete and continuous cues [@szabo2024spontaneous], while Gidgee skinks (_Egernia stokesii_) discriminated between items when they differed in number, but not when size was the main quantitative cue [@szabo_spontaneous_2021]. These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, and that the use of the ANS versus the OFS can differ between species and in different contexts [@szabo_spontaneous_2021]. However, cognitive performance is not solely dictated by species differences; developmental conditions can also shape an individual's ability to process numerical information [@vila_pouca_quantity_2019].

Cognition can be strongly influenced by developmental conditions because the brain is particularly sensitive to environmental inputs in early life stages [@zhu_prenatal_2004]. For instance, the temperature at which embryos develop can affect neuron density and learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. Thermal effects on cognition may arise directly or indirectly through parental effects [@Crino_2023]. For example, temperatures outside the optimal range can trigger stress responses, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. As such, the interaction between GCs and developmental temperature may be relevant in shaping reptile cognition. Most studies on the effects of GC exposure or elevated temperatures during development focus on learning and memory, but few studies have investigated how early-life conditions can affect quantity discrimination despite its importance on animals' lives [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated quantity discrimination in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different number of food items. In addition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect quantity discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Previous studies showed that skink species can discriminate between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different number of crickets even when controlling by other continuous variables like total length and area (i.e. by using the OFS). However, we predicted that numerical discrimination would be affected by early-life conditions. Specifically, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Furthermore, we predicted that increased CORT levels would impair numerical discrimination [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures.

@@

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Egg length and width were measured with digital calipers to the nearest mm (± 0.1 mm), while mass was recorded with a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Eggs were assigned a clutch number and an individual identity. Then we treated eggs with CORT or a vehicle control (see _Manipulating Early Thermal and CORT Environments_ details below) and placed them in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite). We covered the cups with plastic wrap to retain moisture and left the eggs in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see _Manipulating Early Thermal and CORT Environments_ details below) until hatching. Incubators were checked for the presence of hatchlings three times a week.

*Hatchlings* -- Immediately after hatching, we measured snout-vent length (SVL) and tail length (TL) with a ruler to the nearest mm (± 0.1 mm), and mass using a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting and a small water dish. During this period, they were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two different developmental stress treatments: CORT-treated eggs were topically supplied with 5 µL of a CORT solution (10 pg/mL). Control treatment eggs received an equal volume of 100% Ethanol. Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. We selected doses based on previous studies where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs in _Lampropholis delicata_ [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures for this species [@qualls2000post].

The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in a partial split-clutch design. At least one egg per clutch to each treatment, and the remaining eggs were randomly assigned to one of the treatments. When we found less than four eggs in a clutch, we assigned each egg randomly to one of the treatments.

@@

Lizards were acclimatised to the experimental arenas for two weeks before starting the numerical discrimination task (see below). The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). Arenas were placed on seven different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms were the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

@@

The crickets used during the experiment were all the same size, but to control for some continuous traits we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length and total area [see @stancher2015numerical]. We tested if the length occupied by the crickets was similar in both choices (see Supplementary Material - Control of size in both options). Our results showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material - Prey orientation test_).

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. When there was no interaction with any of the choices, both _Latency_, _Choice_, and _Interest_ were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were by one observer (PR), who was blind to the treatment of the lizards, but not the type of test or the number of crickets because the videos were analysed manually.

Motivation plays a key role in animals' performance in spontaneous choice tests [@agrillo2014spontaneous; @bisazza2014experim]. To ensure that lizards remained motivated to forage, they were provided with only one cricket during both acclimatisation and the numerical task (see above). During acclimatisation, lizards were fed at the same time each day as the experiments were conducted. In the numerical task, lizards received a cricket immediately after the test and then fasted for more than 20 hours, which we expected would help standardise fasting levels across individuals. To further control for motivation, we excluded from our analyses any trials in which lizards did not interact with any options or did not eat the cricket provided after the test (n = 0 in both cases). Consequently, we do not consider motivation to be a major factor influencing our results (but see Discussion).

@@

We modelled the three main recorded behaviours: _Latency_, _Choice_, and _Interest_ as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)]. We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included in our predictors sex, and age. We did not find any effect of sex or age on Choice or Interest, but we found an effect on Latency (see Results). Regardless, we included both factors in all models. Age was centered to zero in the model, and we averaged the predicted values for males and females.

We included lizard identity and clutch as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, and so, clutch identity captures potential maternal effects.

We used the posterior distributions of parameters from these models to test for between treatment differences in the ability to discriminate different numbers of crickets for each of the numerical tests. Because the posterior distributions were not always normally distributed, we report the median as the main measure of central tendency. We used the 95% Highest Posterior Density Intervals (95% HPDI) using the hdi function in bayestestR [@bayestestR] to test if the contrasts between treatments for all the variables, or the _Interest_ per each test was different form zero. Given Choice was Bernoulli, we modified our null hypotheses to test if differed from a probability of 0.5 as there were only two possible choices and if animals were choosing randomly we would predict each choice would have a 50% probability.

#### Ethical note
Subjects were collected from a breeding colony that had been kept in the lab since 2019. All animals in this experiment were provided humane laboratory housing, thermoregulation opportunities, and light and humidity conditions (see above for details). Lizards in the colony received five to seven crickets per individual three times a week while the subjects in the experiment were fed once per day. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Upon completion of experimental trials, lizards were kept in their enclosure and will be euthanized following standard humane procedures once they reach the humane endpoint.

>>>>

---
title: "Quantity discrimination, decision-making, and the role of early-life conditions in a lizard"
docx: 
  reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology-and-sociobiology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Abstract
Quantity discrimination affects a range of behaviours essential for fitness, including social interactions, navigation, and foraging. Animals use different cognitive systems to distinguish between quantities. The Approximate Number System (ANS) estimates, while the Object File System (OFS) tracks discrete items. In reptiles, quantity discrimination remained largely unexplored until recently, and the use of one system over the other can depend on the species or the context in which they are tested. Early life conditions can also influence brain development and cognition, potentially affecting quantity discrimination and decision-making. Here, we tested the ability of the common garden skink (_Lampropholis guichenoti_) to discriminate between quantities and the effects of elevated prenatal GCs and incubation temperature on quantity discrimination and decision-making. We assessed quantity discrimination using a spontaneous choice test with food as a stimulus. We subjected lizards to five two-choice tests differing in the number of items (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4). Contrary to our predictions, we found no evidence of quantity discrimination in _L. guichenoti_. Furthermore, prenatal conditions did not influence their performance or decision-making. Despite these results, further investigation of quantity discrimination in live-prey-feeding reptiles and the impact of early conditions on other taxa remains an important area of research.

## Significance statement
Discriminating between quantities can be crucial for individual fitness, yet it remains poorly understood in reptiles. This study investigated quantity discrimination in _Lampropholis guichenoti_ using a spontaneous choice test with food as the stimulus. Contrary to our predictions, we found no evidence of quantity discrimination in this species. In addition, we explored the impact of early environmental factors, specifically prenatal corticosterone exposure and incubation temperature, on quantity discrimination and decision-making. However, we did not observe any effect of these early-life conditions on lizards’ behaviour. These results open new avenues for exploring the evolution of quantity discrimination in reptiles and identifying factors influencing cognitive development in this group. Further research should explore alternative methods, such as trained tasks, to better understand the mechanisms underlying quantity discrimination and decision-making in reptiles.

@@

The main two experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. Spontaneous choice tests ascertain if animals can distinguish between two groups of biologically relevant stimuli that differ quantitatively (number, size, volume, etc) [@agrillo2014spontaneous; @cox2016quantity]. Experiments employing training procedures evaluate an animal’s ability to learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other and can provide different results even in the same species [see @miletto2017quantitative; @miletto2018quantity].

Numerous reasons exist for why methods can lead to different results. First, spontaneous choice tests assume that one of the options presented is more suitable than the other (e.g. the group with more conspecifics or the larger amount of food) when this may not always be the case. For example, choosing groups of conspecifics with fewer individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [@yee2013costs]. As such, context and motivation can be essential drivers of decision-making that can change the way animals discriminate between quantities [@agrillo2014spontaneous]. Second, each method may engage distinct cognitive mechanisms when processing numerical information [@agrillo2014spontaneous]. Different procedures may shape how animals discriminate between quantities, making them more responsive towards certain quantity traits over others. Some tasks may encourage a focus on approximate magnitudes, while others may promote attention to individual items [@agrillo2014spontaneous].

In this regard, researchers propose that two core systems underlie numerical cognition: the approximate number system (ANS) and the object file system (OFS) [@hyde2011two]. The ANS estimates the approximate magnitude of items within a set, with accuracy decreasing as the numerical difference between sets decreases — a principle known as Weber’s law [@hyde2011two]. In contrast, the OFS allows for the precise tracking of discrete items but is thought to be limited to small quantities, typically up to three or four [@hyde2011two]. The engagement of these systems depends not only on the numerical magnitude but also on contextual factors such as the availability of continuous variables (e.g., area, density, movement) and the nature of the task [@stancher2015numerical]. Such continuous cues typically favor ANS-based processing, while in their absence, the OFS is more likely to be employed [@agrillo2014spontaneous], which often complicates the study of numerosity, as continuous traits must be carefully controlled to ensure the use of the OFS [@agrillo2014spontaneous]. In spontaneous choice tests, for instance, animals may rely on cumulative surface area rather than numerical differences when selecting between two sets of food items [@agrillo2014spontaneous]. To disentangle these cues, researchers often keep the total area of the sets constant while changing the number of items, forcing individuals to rely on numerical information alone [@stancher2015numerical].

Quantity discrimination studies have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@mehlis2015quantification; @nieder2018evolution]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in discerning numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) and red-footed turtles (_Chelonoidis carbonaria_) can rapidly learn a quantity discrimination task [@soldati2017long; @lin2021superior; @lin2024trained], and Hermann’s tortoises (_Testudo hermanni_) can discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous]. In lizards, the Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. Shinglebacks (_Tiliqua rugosa_) exhibited spontaneous quantity discrimination using both discrete and continuous cues [@szabo2024spontaneous], while Gidgee skinks (_Egernia stokesii_) discriminated between items when they differed in number but not when size was the main quantitative cue [@szabo_spontaneous_2021]. These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, and that the use of the ANS versus the OFS can differ between species and in different contexts [@szabo_spontaneous_2021]. However, cognitive performance is not solely dictated by species differences; developmental conditions can also shape an individual's ability to process numerical information [@vila_pouca_quantity_2019].

Developmental conditions can strongly influence cognition because the brain is particularly sensitive to environmental inputs in early life stages [@zhu_prenatal_2004]. For instance, the temperature at which embryos develop can affect neuron density and learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. Thermal effects on cognition may arise directly or indirectly through parental effects [@Crino_2023]. For example, temperatures outside the optimal range can trigger stress responses, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. As such, the interaction between GCs and developmental temperature may be relevant in shaping reptile cognition. Most studies on the effects of GC exposure or elevated temperatures during development focus on learning and memory, but few studies have investigated how early-life conditions can affect quantity discrimination despite its importance on animals' lives [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated quantity discrimination in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different numbers of food items. In addition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect quantity discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Previous studies showed that skink species can discriminate between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different numbers of crickets even when controlling by other continuous variables like total length and area (i.e. by using the OFS). However, we predicted that numerical discrimination would be affected by early-life conditions. Specifically, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Furthermore, we predicted that increased CORT levels would impair numerical discrimination [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures.

@@

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Egg length and width were measured with digital calipers to the nearest mm (± 0.1 mm), while mass was recorded with a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Eggs were assigned a clutch number and an individual identity. Then, we treated eggs with CORT or a vehicle control (see _Manipulating Early Thermal and CORT Environments_ details below) and placed them in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite). We covered the cups with plastic wrap to retain moisture and left the eggs in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see _Manipulating Early Thermal and CORT Environments_ details below) until hatching. Incubators were checked for the presence of hatchlings three times a week.

*Hatchlings* -- Immediately after hatching, we measured snout-vent length (SVL) and tail length (TL) with a ruler to the nearest mm (± 0.1 mm), and mass using a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) with non-stick matting and a small water dish. During this period, they were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two developmental stress treatments: CORT-treated eggs were topically supplied with 5 µL of a CORT solution (10 pg/mL). Control treatment eggs received an equal volume of 100% Ethanol. Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. based on previous studies, we selected doses where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs in _Lampropholis delicata_ [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limits of the natural incubation temperatures for this species [@qualls2000post].

The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in a partial split-clutch design. At least one egg per clutch was allocated to each treatment, and the remaining eggs were randomly assigned to one of the treatments. When we found less than four eggs in a clutch, we assigned each egg randomly to one of the treatments.

@@

Lizards were acclimatised to the experimental arenas for two weeks before starting the numerical discrimination task (see below). The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). Arenas were placed on seven different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms being the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

@@

The crickets used during the experiment were all the same size, but to control for some continuous traits, we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length and total area [see @stancher2015numerical]. We tested if the length occupied by the crickets was similar in both choices (see Supplementary Material - Control of size in both options). Our results showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material - Prey orientation test_).

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets from the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. _Latency_, _Choice_, and _Interest_ were noted as NA when there was no interaction with any of the choices. We also considered _Interest_ as NA in a few cases where the animals managed to lift the dish and eat the prey; _Latency_ and _Choice_ were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were by one observer (PR), who was blind to the treatment of the lizards, but not the type of test or the number of crickets because the videos were analysed manually.

Motivation plays a key role in animals' performance in spontaneous choice tests [@agrillo2014spontaneous; @bisazza2014experim]. To ensure that lizards remained motivated to forage, they were provided only one cricket during acclimatisation and the numerical task (see above). During acclimatisation, lizards were fed at the same time each day as the experiments were conducted. In the numerical task, lizards received a cricket immediately after the test and then fasted for more than 20 hours, which we expected would help standardise fasting levels across individuals. To further control for motivation, we excluded from our analyses any trials in which lizards did not interact with any options or did not eat the cricket provided after the test (n = 0 in both cases). Consequently, we do not consider motivation a significant factor influencing our results (but see Discussion).

@@

We modelled the three main recorded behaviours: _Latency_, _Choice_, and _Interest_ as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)]. We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included sex, and age as predictors. We did not find any effect of sex or age on Choice or Interest, but we found an effect on Latency (see Results). Regardless, we included both factors in all models. Age was centered to zero in the model, and we averaged the predicted values for males and females.

We included lizard identity and clutch as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, so clutch identity captures potential maternal effects.

We used the posterior distributions of parameters from these models to test for treatment differences in the ability to discriminate different numbers of crickets for each of the numerical tests. Because the posterior distributions were not always normally distributed, we report the median as the main measure of central tendency. We used the 95% Highest Posterior Density Intervals (95% HPDI) using the hdi function in bayestestR [@bayestestR] to test if the contrasts between treatments for all the variables or the _Interest_ per each test were different from zero. Given Choice was Bernoulli, we modified our null hypotheses to test if it differed from a probability of 0.5 as there were only two possible choices and if animals were choosing randomly we would predict each choice would have a 50% probability.

#### Ethical note
Subjects were collected from a breeding colony that had been kept in the lab since 2019. All animals in this experiment were provided humane laboratory housing, thermoregulation opportunities, and light and humidity conditions (see above for details). Lizards in the colony received five to seven crickets per individual three times a week while the subjects in the experiment were fed once per day. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Upon completion of experimental trials, lizards were kept in their enclosure and will be euthanised following standard humane procedures once they reach the humane endpoint.
